---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Import all libraries to be used
```{r, echo=TRUE}
require(ggplot2)
require(lubridate)
require(gridExtra)
```

## Loading and preprocessing the data

To load the data we just need to extract it from the zip file:
```{r echo=TRUE}
unzip("activity.zip")
```

Then load the csv as follows:
```{r echo=TRUE}
df <- read.csv("activity.csv")
```

Then, we can take an initial look at the data:
```{r echo=TRUE}
head(df, 10)
summary(df)
```


## What is mean total number of steps taken per day?
First, let's extract the daily data into a different data frame:

```{r echo=TRUE}
spd <- data.frame(tapply(df$steps, df$date, FUN = sum))
names(spd) <- c("steps")
head(spd)
summary(spd)
```

Now let's print a histogram of the number of steps per day:
```{r echo=TRUE}
ggplot(spd, aes(x=steps)) + geom_histogram(binwidth = 1000) + geom_vline(aes(xintercept=mean(steps, na.rm = TRUE)), color="maroon") + ggtitle("Steps per day frequency") + xlab("Number of steps") + ylab("Frequency")
```

As you can see the mean and median are as follows:
```{r echo=TRUE}
with(spd, mean(steps, na.rm = TRUE))
with(spd, median(steps, na.rm = TRUE))
```

## What is the average daily activity pattern?
Let's again, prepare a data frame to explore the values:

```{r echo=TRUE}
spi <- data.frame(tapply(df$steps, df$interval, FUN = mean, na.rm=TRUE))
names(spi) <- c("avg")
head(spi)
tail(spi)
summary(spi)
```

Now, let's plot the average steps taken each interval accross the dates:
```{r echo=TRUE}
ggplot(spi, aes(as.numeric(row.names(spi)), avg)) + geom_point() + geom_line() + ggtitle("Average steps at interval") + xlab("Interval") + ylab("Avg. Steps")
```

We can see that the 5 minute interval in which this subject has the maximum number of steps is:
```{r echo=TRUE}
spi[spi$avg == max(spi$avg),]
```

## Imputing missing values
However, the original dataset had lots of NA's that might be changing the result of the analysis. Let's investigate and fix this issue:
```{r echo=TRUE}
length(df$steps)
sum(is.na(df))
sum(is.na(df))/length(df$steps)*100
```

About 13% of the values are missing. Let's replace them by the interval's average:
```{r echo=TRUE}
df$steps2 <- df$steps
df[is.na(df$steps), "steps"] <- spi[,"avg"]
```

Let's create a new dataset with the cleaned values and return the original to the initial state:
```{r echo=TRUE}
df2 <- df
df$steps <- df$steps2
df$steps2 <- NULL
df2$steps2 <- NULL
head(df)
head(df2)
summary(df)
summary(df2)
any(is.na(df$steps))
any(is.na(df2$steps))
```


Now that we took care of missing values, let's calculate the mean and median once again and see the effect of our changes:
```{r echo=TRUE}
spd2 <- data.frame(tapply(df2$steps, df2$date, FUN = sum))
names(spd2) <- c("steps")
head(spd2)
summary(spd2)
```

Now let's print a histogram of the number of steps per day:
```{r echo=TRUE}
ggplot(spd2, aes(x=steps)) + geom_histogram(binwidth = 1000) + geom_vline(aes(xintercept=mean(steps, na.rm = TRUE)), color="maroon") + ggtitle("Steps per day frequency") + xlab("Number of steps") + ylab("Frequency")
```

As you can see the mean and median are as follows:
```{r echo=TRUE}
with(spd2, mean(steps, na.rm = TRUE))
with(spd2, median(steps, na.rm = TRUE))
```

Surprisingly, the median and mean now are equal. This is the effect of changing the NA's values.

## Are there differences in activity patterns between weekdays and weekends?
First, let's add a factor variable representing the weekend and weekdays days:
```{r echo=TRUE}
df2$weekday <- as.factor(ifelse(wday(as.Date(df2$date)) == 1 | wday(as.Date(df2$date)) == 7, "weekend", "weekday"))
summary(df2)
```

Let's extract the values per weekday:
```{r echo=TRUE}
wdd <- subset(df2, df2$weekday == "weekend")
wyd <- subset(df2, df2$weekday == "weekday")
wdspi <- data.frame(tapply(wdd$steps, wdd$interval, FUN = mean, na.rm=TRUE))
wyspi <- data.frame(tapply(wyd$steps, wyd$interval, FUN = mean, na.rm=TRUE))
names(wdspi) <- c("avg")
names(wyspi) <- c("avg")
head(wdspi)
head(wyspi)
tail(wdspi)
tail(wyspi)
summary(wdspi)
summary(wyspi)
```

Now, let's compare the interval activity per weekday:
```{r echo=TRUE}
wdplot <- ggplot(wdspi, aes(as.numeric(row.names(wdspi)), avg)) + geom_point() + geom_line() + ggtitle("Weekend Activity") + xlab("Interval") + ylab("Avg. Steps") 
wyplot <- ggplot(wyspi, aes(as.numeric(row.names(wyspi)), avg)) + geom_point() + geom_line() + ggtitle("Weekday Activity") + xlab("Interval") + ylab("Avg. Steps")
grid.arrange(wdplot, wyplot, nrow=2)
```

Interestingly, the average number of steps in weekdays is greater than in weekends, however, the weekend max is greater than the weekday max.
 
 
 
